Joint maximum likelihood of phylogeny and ancestral states is not consistent

### 2019-03-26 appendix.tex
Doneso.

We begin by showing an inconsistency in branch length estimation on the InvFels tree.

@@

\begin{proof}
Consider a version of $\ell$ given by
\[
\ell_{\tau^*, t^*}(\tau_1, t; \boldsymbol\xi)
\]
where we fix a given value of $\boldsymbol\xi$.
Then, there exists a closed form solution to $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ solving
\[
\hat{t}_{\boldsymbol\xi} = \arg\max_t \ell_{\tau^*, t^*}(\tau_1, t; \boldsymbol\xi).
\]
We show in this case that the log-likelihood $\ell$ attains a unique maximum at $\hat{t}_{\boldsymbol\xi}$.
For fixed $\boldsymbol\xi$, the log-likelihood can be decomposed into a sum of functions of each variable, i.e.,
\begin{align*}
\ell_{\tau^*,t^*}(\tau^*, t; \boldsymbol\xi) &= \sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,x_1}(x_1)+\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,y_1}(y_1)+\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,x_2}(x_2)\\
&\qquad + \sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,y_2}(y_2)+\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,w}(w).
\end{align*}
Due to this additive form, all off-diagonal terms of the Hessian for this function are zero, so we show that the diagonal terms are nonpositive.
For example, if we focus on the variable $x_1$, holding other variables constant, then
\[
\ell(x_1) \sim \sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,x_1}(x_1).
\]
where $\sim$ denotes equality up to an additive constant.
Doing calculation as in Figure~\ref{fig:example_likelihoods}, each functional form, suppressing constants with respect to the focal variable (here $x_1$) and the initial $1/32$ constant, takes the form
\[
h_{j,x_1}(x_1) = (1+x_1)^{e_j}(1-x_1)^{1-e_j}
\]
for $e_j\in\{0,1\}$, which, simplifying, results in
\begin{align}
\ell(x_1) &\sim \left(\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}e_j\right) \log(1+x_1) + \left(\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}(1-e_j)\right) \log (1-x_1)\\
                  &= \left(\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}e_j\right) \log(1+x_1) + \left(1-\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}e_j\right) \log (1-x_1), \label{eq:closed-form-x}
\end{align}
which has second derivative
\[
\ell''(x_1) = -\left(\frac{\sum_j p_{\siteSplit_j}e_j}{(1+x_1)^2}+\frac{1-\sum_j p_{\siteSplit_j}e_j}{(1-x_1)^2}\right).
\]
As $x_1\in(0,1]$, we need only $0 \le \sum_j p_{\siteSplit_j}e_j \le 1$ to imply the diagonal terms of the Hessian are nonpositive.
Since $p_{\siteSplit_j}$ are probabilities, then $\sum_j p_{\siteSplit_j} = 1$.
As $e_j\in\{0,1\}$ by definition, this implies $0 \le \sum_j p_{\siteSplit_j}e_j \le 1$ and $\ell''(x_1) \le 0$.
Applying similar arguments to the other variables, the Hessian for the log-likelihood has nonpositive diagonal terms and off-diagonal terms equal to zero, and $\hat{t}$ uniquely maximizes $\ell$.

>>>>

We begin by showing an inconsistency in branch length estimation on the InvFels tree.
In this proof and other material below, we will use a version of $\ell$ given by
\[
\ell_{\tau^*, t^*}(\tau_1, t; \boldsymbol\xi)
\]
where we fix a given value of the ancestral state split $\boldsymbol\xi$.

@@

\begin{proof}
When $\boldsymbol\xi$ is fixed, there exists a closed form solution to $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ solving
\[
\hat{t}_{\boldsymbol\xi} = \arg\max_t \ell_{\tau^*, t^*}(\tau_1, t; \boldsymbol\xi).
\]
We first show that this maximum is unique.
For fixed $\boldsymbol\xi$, the log-likelihood can be decomposed into a sum of functions of each variable, i.e.,
\begin{align*}
\ell_{\tau^*,t^*}(\tau^*, t; \boldsymbol\xi) &= \sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,x_1}(x_1)+\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,y_1}(y_1)+\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,x_2}(x_2)\\
&\qquad + \sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,y_2}(y_2)+\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,w}(w).
\end{align*}
Due to this additive form, all off-diagonal terms of the Hessian for this function are zero, so we show that the diagonal terms are nonpositive.
For example, if we focus on the variable $x_1$, holding other variables constant, then
\[
\ell(x_1) \sim \sum_{j=1}^\nSiteSplits p_{\siteSplit_j}\cdot\log h_{j,x_1}(x_1).
\]
where $\sim$ denotes equality up to an additive constant.
Doing calculation as in Figure~\ref{fig:example_likelihoods}, each functional form, suppressing constants with respect to the focal variable (here $x_1$) and the initial $1/32$ constant, takes the form
\[
h_{j,x_1}(x_1) = (1+x_1)^{e_j}(1-x_1)^{1-e_j}
\]
for $e_j\in\{0,1\}$, which, simplifying, results in
\begin{align}
\ell(x_1) &\sim \left(\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}e_j\right) \log(1+x_1) + \left(\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}(1-e_j)\right) \log (1-x_1)\\
                  &= \left(\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}e_j\right) \log(1+x_1) + \left(1-\sum_{j=1}^\nSiteSplits p_{\siteSplit_j}e_j\right) \log (1-x_1), \label{eq:closed-form-x}
\end{align}
which has second derivative
\[
\ell''(x_1) = -\left(\frac{\sum_j p_{\siteSplit_j}e_j}{(1+x_1)^2}+\frac{1-\sum_j p_{\siteSplit_j}e_j}{(1-x_1)^2}\right).
\]
As $x_1\in(0,1]$, we need only $0 \le \sum_j p_{\siteSplit_j}e_j \le 1$ to imply the diagonal terms of the Hessian are nonpositive.
Since $p_{\siteSplit_j}$ are probabilities, then $\sum_j p_{\siteSplit_j} = 1$.
As $e_j\in\{0,1\}$ by definition, this implies $0 \le \sum_j p_{\siteSplit_j}e_j \le 1$ and $\ell''(x_1) \le 0$.
Applying similar arguments to the other variables, the Hessian for the log-likelihood has nonpositive diagonal terms and off-diagonal terms equal to zero, and $\hat{t}$ uniquely maximizes $\ell$.

### 2019-03-26 joint_inf.tex
Doneso.

An estimator is said to be statistically consistent if it converges to the generating model with probability one in the large-data limit; existing consistency proofs for maximum likelihood phylogenetics \citep{Allman2008-wd,Chai2011-ff,RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood---this is the correct approach to maximum likelihood, as we are optimizing parameters given only observed data.
These proofs do not readily extend to include estimating ancestral states.
Moreover, examples of inconsistency arising from problems where the number of parameters increases with the amount of data \citep{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
In this case those additional parameters are the states of ancestral sequences.
%page 2 section 2.1 at the top of the second column first paragraph second sentence
Although \citet{Sagulenko2018-xl} explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference.
One would like to know when this approximate technique breaks down, even when sequence data is ``perfect,'' i.e., is generated without sampling error according to the exact model used for inference.

@@

One can also use the possible cases for the form of the joint likelihood (Table~\ref{tab:likelihoods}) to understand the outcome of joint inference for various regions in parameter space.
For example, one can plot the regions in which each of these ancestral state splits attain the maximum (Fig.~\ref{fig:max-anc-state}).
In order to understand the optimal internal branch length in each of these cases, we optimize branch lengths numerically to obtain $\hat{w}$ under joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
Using either technique, we see that the region of inconsistency encompasses almost half of the branch fidelity space: given the correct topology, there are many situations where the interior branch length is estimated to be zero.

\begin{figure}
\centering
%file: plotting/_output/invfels/joint_empirical/non-bias/plot.svg
\includegraphics[width=\textwidth]{empirical-estimate}
\caption{
    Estimates for $\hat{p}_w=(1-\hat{w})/2$, the optimal probability of substitution along the inner edge for the joint objective function \eqref{eq:profile_likelihood}, where the true value for $p_w$ is $p_{y^*}$.
    Regions derived in terms of probabilities of a character change along a branch for ``perfect''data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}a).
    The white region in the lower right highlights which values of $x^*$ and $y^*$ result in an interior branch being estimated as length zero, resulting in an inconsistency.
}
\label{fig:bl-general-inconsistency}
\end{figure}

In estimating the interior branch length $w$, we find a systematic bias in the joint inference procedure even when the true branches are short (Fig.~\ref{fig:empirical-bias}).
As data are generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$.
There are discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}) due to the choice of which ancestral state splits are maximal, so we investigate the bias in the region where $p_{x^*}$ and $p_{y^*}$ are both small, i.e., $p_{x^*}, p_{y^*} \le .1$, as these short-branch cases should be the best settings for joint optimization \citep{Sagulenko2018-xl}.
Although the estimates for $\hat{p}_w$ are better than the estimates when $p_{y^*}$ is small and $p_{x^*}$ is large (Fig.~\ref{fig:bl-general-inconsistency}), joint inference still predictably underestimates the interior branch length.
Additionally, the bias estimates $\hat{p}_w-p_{y^*}$ given $p_{x^*}, p_{y^*} \le .1$ are comparable to the branch lengths (Fig.~\ref{fig:empirical-bias}), showing that even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.
One can see the region in which the internal branch length is estimated to be zero under the curve in the lower right of this plot.
Dr.\ Neher has pointed out in his review that the shape of this region can be intuitively understood as that where two mutations on the longer pendant branches are more likely ($p_{x^*}^2$) than one mutation on the internal branch ($p_{y^*}$).

@@

On a practical note, we would like to emphasize that TreeTime implements marginal mode inference.
In January 2019 a warning was added, in all caps, that is triggered when joint inference is used for trees such that any branch length is greater than 0.15.

\section*{Acknowledgements}
We thank Richard Neher, Vladimir Minin, and Joe Felsenstein for helpful discussions.
We are especially thankful to Dr.\ Neher for his helpful suggestions and for providing TreeTime simulation code that verified our conclusions on simulated data.
Insight from the reviewers and the editors greatly improved this manuscript.

>>>>

An estimator is said to be statistically consistent if it converges to the generating model with probability one in the large-data limit; existing consistency proofs for maximum likelihood phylogenetics \citep{Allman2008-wd,Chai2011-ff,RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood---this is the correct approach to maximum likelihood, as it optimizes parameters given only observed data.
These proofs do not readily extend to include estimating ancestral states.
Moreover, examples of inconsistency arising from problems where the number of parameters increases with the amount of data \citep{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
In this case those additional parameters are the states of ancestral sequences.
%page 2 section 2.1 at the top of the second column first paragraph second sentence
Although \citet{Sagulenko2018-xl} explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference.
One would like to know when this approximate technique breaks down, even when sequence data is ``perfect,'' i.e., is generated without sampling error according to the exact model used for inference.

@@

One can also use the possible cases for the form of the joint likelihood (Table~\ref{tab:likelihoods}) to understand the outcome of joint inference for various regions in parameter space.
For example, one can plot the regions in which each of these ancestral state splits attain the maximum (Fig.~\ref{fig:max-anc-state}).
In order to understand the optimal internal branch length in each of these cases, we optimize branch lengths numerically to obtain $\hat{w}$ under joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
We find that for almost half of branch fidelity space, the interior branch length is estimated to be zero given the correct topology.

\begin{figure}
\centering
%file: plotting/_output/invfels/joint_empirical/non-bias/plot.svg
\includegraphics[width=\textwidth]{empirical-estimate}
\caption{
    Estimates for $\hat{p}_w=(1-\hat{w})/2$, the optimal probability of substitution along the inner edge for the joint objective function \eqref{eq:profile_likelihood}, where the true value for $p_w$ is $p_{y^*}$.
    Regions derived in terms of probabilities of a character change along a branch for ``perfect''data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}a).
    The white region in the lower right highlights which values of $x^*$ and $y^*$ result in an interior branch being estimated as length zero.
}
\label{fig:bl-general-inconsistency}
\end{figure}

In estimating the interior branch length $w$, we find a systematic bias in the joint inference procedure even when the true branches are short (Fig.~\ref{fig:empirical-bias}).
As data are generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$.
There are discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}) due to the choice of which ancestral state splits are maximal, so we investigate the bias in the region where $p_{x^*}$ and $p_{y^*}$ are both small, i.e., $p_{x^*}, p_{y^*} \le .1$, as these short-branch cases should be the best settings for joint optimization \citep{Sagulenko2018-xl}.
Although the estimates for $\hat{p}_w$ are better than the estimates when $p_{y^*}$ is small and $p_{x^*}$ is large (Fig.~\ref{fig:bl-general-inconsistency}), joint inference still predictably underestimates the interior branch length.
Additionally, the bias estimates $\hat{p}_w-p_{y^*}$ given $p_{x^*}, p_{y^*} \le .1$ are comparable to the branch lengths (Fig.~\ref{fig:empirical-bias}), showing that even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.
One can see the region in which the internal branch length is estimated to be zero under the curve in the lower right of this plot.
Dr.\ Neher has pointed out during peer review that the shape of this region can be intuitively understood as that where two mutations on the longer pendant branches are more likely (order $p_{x^*}^2$) than one mutation on the internal branch (order $p_{y^*}$).

@@

On a practical note, we would like to emphasize that TreeTime also implements marginal mode inference, which calculates the classical integrated likelihood.
In January 2019 a warning was added, which cannot be turned off and is in all capital letters, that is triggered when joint inference is used for trees such that any branch length is greater than 0.15.

\section*{Acknowledgements}
We thank Richard Neher, Vladimir Minin, and Joe Felsenstein for helpful discussions.
We are especially thankful to Dr.\ Neher for his suggestions and for providing TreeTime simulation code that verified our conclusions on simulated data.
Insight from all of the reviewers and the editors greatly improved this manuscript.
